STA610 Case Study 1 Team 4 Report

### 2021-03-21 final_report.Rmd
Added Lauren's EDA

```{r, echo = FALSE}

load("~/Documents/Spring 2021/STA 610 - H Models/cs1/streetrx.RData")
df <- streetrx %>% filter(api_temp=="oxymorphone", !is.na(ppm))
```

@@

```{r, echo = FALSE}
# cleaning state:

# table(df$state)

# remove levels corresponding to states with 0 observations in df
df$state <- droplevels(df$state)
# table(df$state)

# drop North Dakota - one observation
df <- df %>% filter(state != "North Dakota")
df$state <- droplevels(df$state)
# table(df$state)

# replace state "USA" with "Unknown"
df$state <- recode_factor(df$state, "USA" = "Unknown")
# table(df$state)

### PUT IN APPENDIX: the state-by-region table of coutns below
# table(df$state, df$USA_region)
```

@@

```{r, echo = FALSE}
# Change date to numeric (easier for interpretation)

# this sets 
df$date_num <- as.numeric(as.Date(df$price_date, "%m/%d/%y")) - 14621 # sets 1/12/2010 (first post-2010 date) as zero

tempdf <- df %>% select(date_num, price_date)

# this filters out all data before the 1/1/2000, which probably shouldn't be included
df <- df %>% filter(date_num >= 0)
```

@@

```{r, echo = FALSE}
# recode sources

# remove unused levels
df$source <- droplevels(df$source)

source_df <- data.frame(table(df$source))

# levels(df$source)
levels(df$source)[1] <- "Blank"

source_df <- data.frame(table(df$source))

df$source <- recode_factor(df$source, "Blank" = "Blank",
                           "Internet" = "Internet",
                           "Internet Pharmacy" = "Internet Pharmacy",
                           "Personal" = "Personal",
                           "Heard it" = "Heard it",
                           "Idk" = "Unknown",
                           "w" = "Unknown",
                           "STREET PRICE" = "Heard it",
                           .default = "Internet") # everything else is a URL

source_df_2 <- data.frame(table(df$source))

# drop unknowns

df <- df %>% filter(df$source != "Unknown")
df$source <- droplevels(df$source)

knitr::kable(table(df$source), col.names = c("Level", "Freq"))
```

@@

```{r}
# remove unused levels
df$Primary_Reason <- droplevels(df$Primary_Reason)

reason_df <- data.frame(table(df$Primary_Reason))

# levels(df$Primary_Reason)
levels(df$Primary_Reason)[1] <- "Blank"

reason_df <- data.frame(table(df$Primary_Reason))

df$Primary_Reason <- recode_factor(df$Primary_Reason, "Blank" = "Blank",
                                   "0 Reporter did not answer the question" = "Left unanswered",
                                   "Other or unknown" = "Other or unknown",
                                   "9 To self-treat my pain" = "Self-treat",
                                   "3 To prevent or treat withdrawal" = "Self-treat",
                                   "10 To treat a medical condition other than pain" = "Self-treat",
                                   "4 For enjoyment/to get high" = "Enjoyment",
                                   "5 To resell" = "Resell",
                                   "8 Prefer not to answer" = "Refuse to answer",
                           .default = "Other or unknown")

reason_df_2 <- data.frame(table(df$Primary_Reason))

#print(reason_df)
knitr::kable(reason_df_2, col.names = c("Level", "Freq"))
```

@@

```{r, fig.height=3, fig.align='center'}
# log transform ppm
df$log_ppm <- log(df$ppm)

hist_ppm <- ggplot(df, aes(x=ppm)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", bins = 50)+
 geom_density(alpha=.2, fill="red")

hist_log_ppm <- ggplot(df, aes(x=log_ppm)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", bins = 50)+
 geom_density(alpha=.2, fill="blue")

grid.arrange(hist_ppm, hist_log_ppm, ncol=2)
```

@@

```{r, fig.align='center'}
fl_lppmbycity <- ggplot(df[df$state=="Florida",], aes(x=log_ppm, y=reorder(city, city, length))) +
geom_boxplot() +
labs(title="log PPM by city (FL)", x="Log PPM",y="city")

ca_lppmbycity <- ggplot(df[df$state=="California",], aes(x=reorder(city, city, length), y=log_ppm)) +
geom_boxplot() +
labs(title="log PPM by city (CA)", x="Log PPM",y="city")

grid.arrange(fl_lppmbycity, ca_lppmbycity, ncol= 2)
```

@@

```{r}
group_int_list <- NULL 
group_slope_list <- NULL

for(statej in
  unique(df$state))
{
m <- lm(data = df[df$state==statej, ], log_ppm ~ dose_per_mg)

group_int_list <- c(group_int_list, coef(m)[1])
group_slope_list <- c(group_slope_list, coef(m)[2])
}

hist(group_int_list)
hist(group_slope_list)
```

@@

```{r}
group_int_list <- NULL 
group_slope_list <- NULL

for(statej in
  unique(df$state))
{
m <- lm(data = df[df$state==statej, ], log_ppm ~ date_num)

group_int_list <- c(group_int_list, coef(m)[1])
group_slope_list <- c(group_slope_list, coef(m)[2])
}

hist(group_int_list)
hist(group_slope_list)
```

# Model selection and diagnostics ---- Unfinished! ------

## Model specification

We are considering models of the following form (with some coefficients potentially equal to zero):

$$
y_{ij} = \beta_{0,j} + \beta_{1,j} date\text{_}num + \beta_{2,j} mgstr + \beta_{3,j} bulk\text{_}purchase + \beta_{4,j} Primary\text{_}Reason + \beta_{5,j} source + \beta_{6,j} date\text{_}quarter + \epsilon_{ij}
$$

Where: 
* $\beta_{0,j}$ represents the intercept, which we plan to fit as a random intercept for each group $j$, so that it would have the form:
$\beta_{0,j} = \beta_0 + b_{0,j}, b_{0,j} \sim N(0,\tau_0^2)$

* The $\beta_k$ for $k=1,...,6$ are the coefficients for our fixed effect predictors. We will consider random slopes by group for these, in which case they would have the form: 
$\beta_{k,j} = \beta_k + b_{k,j}, b_{k,j} \sim N(0,\tau_k^2)$

We also have that $\epsilon_{ij} \perp b_{0,j} \overset{iid}{\sim} N(0,\sigma^2)$ for all $i, j$.

## Model specification

### Grouping variables and random intercepts

We see that `state` appears to be best for grouping; region is good as well.

```{r}
mod_s <- lmer(data=df, log_ppm ~ (1 |state) + date_num + date_quarter + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F)
mod_c <- lmer(data=df, log_ppm ~ (1 |city) + date_num + date_quarter + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F)
mod_r <- lmer(data=df, log_ppm ~ (1 |USA_region) + date_num + date_quarter + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F)

summary(mod_s)
# summary(mod_c)
# summary(mod_r)

LR <- 2*(logLik(mod_s)-logLik(mod_c))
0.5*(1-pchisq(LR[1],1)+1-pchisq(LR[1],1)) # choose s over c

LR <- 2*(logLik(mod_r)-logLik(mod_c))
0.5*(1-pchisq(LR[1],1)+1-pchisq(LR[1],1)) # choose c over r

LR <- 2*(logLik(mod_s)-logLik(mod_r))
0.5*(1-pchisq(LR[1],1)+1-pchisq(LR[1],1)) # choose s over r

# conclusion: choose state random intercept
```

```{r}
anova(mod_s,
      lmer(data=df, log_ppm ~ (1 |state) + date_quarter + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F))
# drop date_num

anova(lmer(data=df, log_ppm ~ (1 |state) + date_quarter + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F),
      lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F))
# drop date_quarter

anova(lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F),
      lmer(data=df, log_ppm ~ (1 |state) + bulk_purchase + Primary_Reason + source, REML=F))
# don't drop dose_per_mg

anova(lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F),
      lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + Primary_Reason + source, REML=F))
# don't srop bulk_purchase

anova(lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase + Primary_Reason + source, REML=F),
      lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase + source, REML=F))
# do drop Primary_reason

anova(lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase + source, REML=F),
      lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase, REML=F))
# do drop source
```

Now investigate random slopes:

```{r}
anova(lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase, REML=F),
      lmer(data=df, log_ppm ~ (dose_per_mg |state) + dose_per_mg + bulk_purchase, REML=F))
# random slopes for dose_per_mg are good - BIC = 8624.6

anova(lmer(data=df, log_ppm ~ (1 |state) + dose_per_mg + bulk_purchase, REML=F),
      lmer(data=df, log_ppm ~ (bulk_purchase |state) + dose_per_mg + bulk_purchase, REML=F))
# random slopes for nulk_purchase not helpful
```

Our candidate for final model is then:

```{r}
mod_final <- lmer(data=df, log_ppm ~ (dose_per_mg |state) + dose_per_mg + bulk_purchase, REML=F)
```

There are convergence inssues. Try removing states that are influential? Based on the "dfbetas" visualization below, they are Wyoming, Nevada, Alaska.

```{r}
mod_infl <- influence(mod_final,"state")
print(2/sqrt(length(unique(df_filter$state))))
diagn_df <- data.frame(dfbetas(mod_infl))
plot(mod_infl,which="dfbetas",xlab="DFBETAS",ylab="State")

resqq <- plot_resqq(mod_final)
ranefci <- plot_ranef(mod_final)
grid.arrange(resqq, ranefci, ncol = 2)
```

```{r}
df_filter <- df %>% filter(state != "Alaska" & state != "Wyoming" & state != "Nevada")
```

```{r}
mod_final <- lmer(data=df_filter, log_ppm ~ (dose_per_mg |state) + dose_per_mg + bulk_purchase, REML=F)

mod_infl <- influence(mod_final,"state")
print(2/sqrt(length(unique(df_filter$state))))
diagn_df <- data.frame(dfbetas(mod_infl))
plot(mod_infl,which="dfbetas",xlab="DFBETAS",ylab="State")

resqq <- plot_resqq(mod_final)
ranefci <- plot_ranef(mod_final)
grid.arrange(resqq, ranefci, ncol = 2)
```

See below: looks like the singularity is because some states don't have enough data for `bulk_purchase`. Ugh. Should we use `region` instead???

```{r}
View(data.frame(table(df$state, df$bulk_purchase)))
```

>>>>

```{r, echo = FALSE}
load("~/Documents/Spring 2021/STA 610 - H Models/cs1/streetrx.RData")
df <- streetrx %>% filter(api_temp=="oxymorphone", !is.na(ppm))
```

@@

```{r, echo = FALSE}
# cleaning state:
# table(df$state)
# remove levels corresponding to states with 0 observations in df
df$state <- droplevels(df$state)
# table(df$state)
# drop North Dakota - one observation
df <- df %>% filter(state != "North Dakota")
df$state <- droplevels(df$state)
# table(df$state)
# replace state "USA" with "Unknown"
df$state <- recode_factor(df$state, "USA" = "Unknown")
# table(df$state)
### PUT IN APPENDIX: the state-by-region table of coutns below
# table(df$state, df$USA_region)
```

@@

```{r, echo = FALSE}
# Change date to numeric (easier for interpretation)
# this sets 
df$date_num <- as.numeric(as.Date(df$price_date, "%m/%d/%y")) - 14621 # sets 1/12/2010 (first post-2010 date) as zero
tempdf <- df %>% select(date_num, price_date)
# this filters out all data before the 1/1/2000, which probably shouldn't be included
df <- df %>% filter(date_num >= 0)
```

@@

```{r, echo = FALSE}
# recode sources 
# remove unused levels
df$source <- droplevels(df$source)
source_df <- data.frame(table(df$source))
# levels(df$source)
levels(df$source)[1] <- "Blank"
source_df <- data.frame(table(df$source))
df$source <- recode_factor(df$source, "Blank" = "Blank",
                           "Internet" = "Internet",
                           "Internet Pharmacy" = "Internet Pharmacy",
                           "Personal" = "Personal",
                           "Heard it" = "Heard it",
                           "Idk" = "Unknown",
                           "w" = "Unknown",
                           "STREET PRICE" = "Heard it",
                           .default = "Internet") # everything else is a URL
source_df_2 <- data.frame(table(df$source))
# drop unknowns
df <- df %>% filter(df$source != "Unknown")
df$source <- droplevels(df$source)
knitr::kable(table(df$source), col.names = c("Level", "Freq"))
```

@@

```{r}
# remove unused levels
df$Primary_Reason <- droplevels(df$Primary_Reason)
reason_df <- data.frame(table(df$Primary_Reason))
# levels(df$Primary_Reason)
levels(df$Primary_Reason)[1] <- "Blank"
reason_df <- data.frame(table(df$Primary_Reason))
df$Primary_Reason <- recode_factor(df$Primary_Reason, "Blank" = "Blank",
                                   "0 Reporter did not answer the question" = "Left unanswered",
                                   "Other or unknown" = "Other or unknown",
                                   "9 To self-treat my pain" = "Self-treat",
                                   "3 To prevent or treat withdrawal" = "Self-treat",
                                   "10 To treat a medical condition other than pain" = "Self-treat",
                                   "4 For enjoyment/to get high" = "Enjoyment",
                                   "5 To resell" = "Resell",
                                   "8 Prefer not to answer" = "Refuse to answer",
                           .default = "Other or unknown")
reason_df_2 <- data.frame(table(df$Primary_Reason))
#print(reason_df)
knitr::kable(reason_df_2, col.names = c("Level", "Freq"))
```

@@

```{r, fig.height=3, fig.align='center'}
# log transform ppm
df$log_ppm <- log(df$ppm)
hist_ppm <- ggplot(df, aes(x=ppm)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", bins = 50)+
 geom_density(alpha=.2, fill="red") 
hist_log_ppm <- ggplot(df, aes(x=log_ppm)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", bins = 50)+
 geom_density(alpha=.2, fill="blue") 
grid.arrange(hist_ppm, hist_log_ppm, ncol=2)
```

@@

```{r, fig.align='center'}
fl_lppmbycity <- ggplot(df[df$state=="Florida",], aes(x=log_ppm, y=reorder(city, city, length))) +
geom_boxplot() +
labs(title="log PPM by city (FL)", x="Log PPM",y="city")
ca_lppmbycity <- ggplot(df[df$state=="California",], aes(x=reorder(city, city, length), y=log_ppm)) +
geom_boxplot() +
labs(title="log PPM by city (CA)", x="Log PPM",y="city")
grid.arrange(fl_lppmbycity, ca_lppmbycity, ncol= 2)
```

@@

```{r}
group_int_list <- NULL 
group_slope_list <- NULL
for(statej in
  unique(df$state))
{
m <- lm(data = df[df$state==statej, ], log_ppm ~ dose_per_mg)
group_int_list <- c(group_int_list, coef(m)[1])
group_slope_list <- c(group_slope_list, coef(m)[2])
}
hist(group_int_list)
hist(group_slope_list)
```

@@

```{r}
group_int_list <- NULL 
group_slope_list <- NULL
for(statej in
  unique(df$state))
{
m <- lm(data = df[df$state==statej, ], log_ppm ~ date_num)
group_int_list <- c(group_int_list, coef(m)[1])
group_slope_list <- c(group_slope_list, coef(m)[2])
}
hist(group_int_list)
hist(group_slope_list)
```
